MLH1 Results Outline

### 2020-02-13 doc/Results.rmd
cleaning up more

---
title: "MLH1 Results Outline"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
bibliography: refs_Main_Report.bib
---

@@

@@

@@

```{r first.boxplots, echo=FALSE, fig.align='center', warning=FALSE}

#the below ordering works!! keep it
MLH1_data$strain<- factor(MLH1_data$strain, levels =c("G", "LEW","WSB",
              "PWD","CZECH","SKIVE",
              "KAZ","TOM","AST",
              "MSM","MOLF",
              "CAST","HMI","SPRET","SPIC", "other"), order=T )
#order the data frame
MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

justMuscu <- ggplot(MLH1_data[MLH1_data$species == "M.musculus", ])+
  geom_boxplot(aes(x = as.factor(strain), y = adj_nMLH1.foci, fill= strain,alpha=0.5) )+ ggtitle("Boxplots of MLH1 distributions") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
 ylim( c((min(MLH1_data$adj_nMLH1.foci)-3) ), (max(MLH1_data$adj_nMLH1.foci)+3)) +   scale_fill_manual(values=colors_of_strains)
#facet_grid(~sex, scales="free_x")+scale_color_manual(values=colors_of_strains)+

justMuscu <- justMuscu   + facet_wrap(~ sex, scales="free")

#scales="free_y"
#this is ok except for the set placement of mice on x axis
#if the full dataset was ordered correctly, this would kinda fix it

#boxplots by strain
ss <- ggplot(MLH1_data, aes(x = as.factor(strain), y = adj_nMLH1.foci, fill= strain )) + geom_boxplot(aes(alpha=0.5))  + ggtitle("Boxplots of MLH1 distributions by strain") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  ylim( c( (min(MLH1_data$adj_nMLH1.foci)-3), (max(MLH1_data$adj_nMLH1.foci)+5)) )+scale_fill_manual(values=colors_of_strains)

ss <- ss   + facet_wrap(~ sex, scales="free")

#these plot currently look bad
```

@@

#this move this to the setup file

strain.table.HQ_Fig.og <- ddply(MLH1_data, c("strain", "sex"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  =length(nMLH1.foci),
                        mean_co =as.numeric(format(round(mean(nMLH1.foci), 3 ), nsmall=3) ),
                        cV = cv(nMLH1.foci, na.rm = TRUE),
                        var = as.numeric(format(round(   var(nMLH1.foci, na.rm = TRUE),3), nsmall=3)), 
                        sd   = round(sd(nMLH1.foci, na.rm = TRUE), 3),
                       #Standard error (mice vs cells)
                        se  = round(sd / sqrt(Ncells), 3)
    )

@@

```{r maw.fig1s, echo=FALSE, include=FALSE}

@@

#table for the means by strain

table( droplevels(female_mouse.avs_for.analysis$strain) )
#there are cast strains too

musc.female.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))
musc.female.plot

Mol.female.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,33))
Mol.female.plot

Dom.female.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,33))
Dom.female.plot
#Gough female is higher

#female strain table
#female_mouse.avs_for.analysis
F.strain.table <- ddply(female_mouse.avs_for.analysis, c("strain"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  = sum(Ncells),
                        mean_co =as.numeric(format(round(mean(mean_co), 3 ), nsmall=3) ),

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se)
    )

G and LEW have a higher strain means compared to WSB. MSM (and PWD) also has an elevated mean.

The variance across mouse means for each strain is also variable across strains.

In the dom  subspecies the ranges of mouse means are `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co  )` in WSB, `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co  )` in G, `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co  )` in LEW.

In the musc subspecies the ranges of mouse means are `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "PWD"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "PWD"))$mean_co  )` in PWD, `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "KAZ"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "KAZ"))$mean_co  )` in KAZ.

The range is `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co  )` in MSM.

Two glm models were run to test the effects of strain and subspecies.

glm.female.strain.subsp <- lm(mean_co ~ subsp * strain, data= female_mouse.avs_for.analysis)

summry.female.strain.effects <- summary(glm.female.strain.subsp)

round(summry.female.strain.effects$coefficients[,c(1,4)],5)

#write.table(summry.female.strain.effects$coefficients, "~./MLH1repo/summry.female.strain.effects.csv", sep=",",row.names = TRUE)

#how much greater is G female
G_female <-  female_mouse.avs_for.analysis %>% filter(strain == "G")
mean.G.female <- mean(G_female$mean_co)

sans_high_female <- female_mouse.avs_for.analysis %>% filter(strain != "G") %>%  filter(strain != "MSM") %>%  filter(strain != "LEW")
mean.all.other <- mean(sans_high_female$mean_co)

#sans_MSM_female / mean.all.other

## within mouse variance models 
lmVAR.female.strain.subsp <- lm(var ~ subsp * strain, data= female_mouse.avs_for.analysis)
summry.female.strain.VAReffects <- summary(lmVAR.female.strain.subsp)

lmVAR.female.strain <- lm(var ~ strain, data= female_mouse.avs_for.analysis)
lmVAR.female.strain_sum <- summary(lmVAR.female.strain)

lm.m1.Q12_female <- lm(var ~ subsp * strain, data= Q12_female_table)
lm.m1.Q12_female_sum <- summary(lm.m1.Q12_female)
#LEW not sig

lm.m2.Q12_female <- lm(var ~ strain, data= Q12_female_table)
lm.m2.Q12VAR.female_sum <- summary(lmVAR.female.strain)
#LEW sig
```

G is `r mean.G.female / mean.all.other` higher the the other means. LEW is `r mean.LEW.female / mean.all.other` higher  and MSM is `r mean.MSM.female / mean.all.other`. These three will be designated as 'moderate high rec' strains for later sex specific analysis.

The male specific analysis was done to assess the variance across strains. The plots below illustrate mouse level means for MLH1 per cell serperated by subspecies.

Dom.male.plot <- ggplot(Dom.male_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom male of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,33))
Dom.male.plot

#female strain table
#female_mouse.avs_for.analysis
M.strain.table <- ddply(just.male_mouse.av, c("strain"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  = sum(Ncells),
                        mean_co =as.numeric(format(round(mean(mean_co), 3 ), nsmall=3) ),

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se)
    )

summry.male.strain.effects <- summary(glm.male.strain.subsp)

SKIVE_male <-   just.male_mouse.av %>% filter(strain == "SKIVE")
#mean(SKIVE_male$mean_co) /

## within mouse variance models 
lmVAR.male.strain.subsp <- lm(var ~ subsp * strain, data= just.male_mouse.av)
summry.male.strain.VAReffects <- summary(lmVAR.male.strain.subsp)
#all p values kinda high, MSM and TOM are slightly sig

lmVAR.male.strain <- lm(var ~ strain, data= just.male_mouse.av)
lmVAR.male.strain_sum <- summary(lmVAR.male.strain)
#non sig

lmcv.male.strain <- lm(cV ~ strain, data= just.male_mouse.av)
lmcv.male.strain_sum <- summary(lmcv.male.strain)
#HMI sig for cv model -- in the other model, CAST subsp and TOM strain are slightly significant

lm.m1.Q12_male <- lm(var ~ subsp * strain, data= Q12_male_table)
lm.m1.Q12_male_sum <- summary(lm.m1.Q12_male)
#NS

lm.m2.Q12_male <- lm(var ~ strain, data= Q12_male_table)
lm.m2.Q12VAR.male_sum <- summary(lm.m2.Q12_male)
#NS

The strain average for PWD, MSM and SKIVE are  ` r mean(PWD_male$mean_co) /mean(sans_LewRec_male$mean_co)`, `r mean(MSM_male$mean_co) / mean(sans_LewRec_male$mean_co)`, and `r mean(SKIVE_male$mean_co) / mean(sans_LewRec_male$mean_co)` higher than the other strain means respectively. Due to this and the significant strain effects they will be designedated into the high rec group for later analyses.

@@

In order to understand the variance within an evolutionary framework we fit the mouse gwRRs to a mixed model, which specified subspecies, sex and the interaction as fixed effects. The subspecies effect - is a proxy for measuring the degree of divergence across subspecies. The strains were coded as a random effect to approximate standing genetic variation across subspecies ranges.

These are the basic Mixed Models we used as a frame work to analyze all the MLH1 count data.

$$mouse \ av.\  metric ~=~ subsp * sex + rand(strain) + \varepsilon $$

The mixed model lets us separate effects to see how they effect a mouse mean CO count. The subspecies effect is a proxy for **Divergence** and the random strain effect is a proxy for **Polymorphism**.

<!--
Setting up the data set for using the mixed model (base data set, HQ data set)

subspecies levels: 3, Musc, Dom and Molossinus

strain levels: Dom: 3, Musc: 3  Mol: 2

sex: 2 female male  
-->

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE, eval=TRUE}
#DF.HetC.MixedModel.HQ

#SWITCH TO HQ.DF
#mean_co.HQ  (this might remove some of the MOLF males)
#DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[DF.HetC.MixedModel$Ncells > 9,]  #this removes the MOLF female

#table(DF.HetC.MixedModel$strain, DF.HetC.MixedModel$sex)

#afriad this is removing NA's
#DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[( (DF.HetC.MixedModel$age.weeks < 25) &  #(DF.HetC.MixedModel$age.days != 13) ) | is.na(DF.HetC.MixedModel$age.weeks), ] #keeps mice without ages, and removes juvinilles

#table(DF.HetC.MixedModel.HQ$strain, DF.HetC.MixedModel.HQ$sex)

#add_subsp function re-writen

#mean_co.HQ.mol
#DF.HetC.MixedModel.HQ$subsp <- ifelse(grepl("MSM", DF.HetC.MixedModel.HQ$strain), "Mol", 
#        ifelse(grepl("MOLF", DF.HetC.MixedModel.HQ$strain), "Mol", DF.HetC.MixedModel.HQ$subsp))

# For some reason I changed subsp to nums (wtf)
#DF.HetC.MixedModel.HQ$subsp <- ifelse(grepl(1, DF.HetC.MixedModel.HQ$subsp), "Dom", 
#        ifelse(grepl(2, DF.HetC.MixedModel.HQ$subsp), "Musc", DF.HetC.MixedModel.HQ$subsp))

#DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[!(is.na(DF.HetC.MixedModel.HQ$mouse) | #DF.HetC.MixedModel.HQ$mouse==""), ]

#remove blank rows (which are these?)
#DF.HetC.MixedModel <- DF.HetC.MixedModel[!(is.na(DF.HetC.MixedModel$mouse) | #DF.HetC.MixedModel$mouse==""), ]
#151 mice, no rows with nas for values

This is the break down of mice used within the Mixed model

@@

Allows random effect from strain. Considers the wild derived inbred strains as random samples from each subspecies territory.

@@

#summary(MLH1.MM.M1)

MM.Fixed.pval <- drop1(MLH1.MM.M1, test="Chisq")$`Pr(Chi)`[2]

anova(MLH1.MM.M1, MLH1.MM.M1.reduce.sex)
anova(MLH1.MM.M1, MLH1.MM.M1.reduce.subp)

MM.rand.pval <- exactRLRT(MLH1.MM.M1)$p.value

#as.data.frame(round(summary(MLH1.MM.M1)$coefficients, 6))

#only the interaction term shows up, when the 2 terms are additive, they both show up in the drop1 results

The significance of the interaction effect for the 2 fixed effects is `r MM.Fixed.pval`. The random effect for strain effect is `r MM.rand.pval`.

### glm followup

To follow up on the mixed model results, we also ran glm / linear models to get a better idea of the strain by sex interaction effects.

 $$mouse \ av \ meteric ~=~ subsp * sex * strain + \varepsilon$$

Post hoc model for investigating strain as a fixed effect.

@@

summ <- summary(glm.strain.fixed.full)
#table of just coeff anf pvalue
round(summ$coefficients[,c(1,4)],5 )

#write.table(summ$coefficients , "~./MLH1repo/glm.strain.fixed.full.csv", sep=",",row.names = TRUE)

Above is the table of coefficients for MM.3, HQ data set, lm function, sex, strain and subsp are interacting fixed effects.

Strain G is significant (p=0.0005), (remember female result)

interaction terms of sex and strain for (G, MSM, PWD).

How are these models different from the last one -- MM4 is missing subsp, so only sex * strain effects.

$$mouse \ av \ metric ~=~ sex * strain + \varepsilon$$

@@

#write.table(summ.strain.fixed$coefficients , "~./MLH1repo/summ.strain.fixed.csv", sep=",",row.names = TRUE)

#as.data.frame( round( summary(strain.fixed)$coefficients, 6))

#summary(strain.fixed)
#maybe this WSB effec is just in the HQ data set?
```

<!--
 Contrasts and ANOVA, mean CO count full vs reduced

Use Anova to compare full vs reduced models
the reduced models will be those missing either, strain, sex, or subsp.

```{r anova.full.v.reduced, echo=FALSE, eval=FALSE}

#strain as fixed
#CONTRAST models
#newnew <- lme(mean_co ~ sex * strain, data=DF.HetC.MixedModel )

HetC.MixedModel.mean_co <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel, random=list(strain=pdDiag(~sex) ) )
R.strain.HetC.MixedModel.mean_co <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel, random = ~1|strain)#musc-male fixed effect is now significant, and male

full.Reduced.HetC.MixedModel.mean_co <- gls(mean_co ~ subsp * sex, data=DF.HetC.MixedModel)

#base vs strain
fullVstrain <- anova(HetC.MixedModel.mean_co, R.strain.HetC.MixedModel.mean_co) #<.0001

#show scientific vs normal
#vv <- fullVstrain$`p-value`
#round(vv[2], 8)
#format(vv[2], scientific=F)

#strain vs no random full.reduce
fullVnon <- anova(HetC.MixedModel.mean_co, full.Reduced.HetC.MixedModel.mean_co) #<.0001

strain.vs.non <- anova(R.strain.HetC.MixedModel.mean_co, full.Reduced.HetC.MixedModel.mean_co) #<.0001

#fullreduce vs strain
#can't distinguish which random effects are more significant -- both tests are significant

fullVreduce22 <- anova(HetC.MixedModel.mean_co, full.Reduced.HetC.MixedModel.mean_co)

#anova for comparing reduced and full models -- 1 of the models should be strain as fixed instead
full2strain <- anova(HetC.MixedModel.HQ, Reduced.strain.HQ)#slightly significant

full.v.fixed <- anova(HetC.MixedModel.HQ, strain.fixed)
#this comparison not useful when there are different fixed effects

#full2noRandom <- anova(HetC.MixedModel.HQ, Reduced.all.HQ) # < .0001

#strain2noRandom <- anova(Reduced.strain.HQ, Reduced.all.HQ) #NS

#anova tests
#full vs strain
full2strain.mol <- anova(HetC.MixedModel.mol, Reduced.strain.mol)# < .0001
#full vs non
full2noRandom.mol <- anova(HetC.MixedModel.mol, Reduced.all.mol)# < .0001
#strain vs non
strain2noRandom.mol <- anova(Reduced.strain.mol, Reduced.all.mol)# < .0001

anova(Reduced.strain.mol, Reduced.all.mol, HetC.MixedModel.mol)

#base vs mol model (this only makes sense if these are nested models)
anova(HetC.MixedModel.mean_co, HetC.MixedModel.mol)
```

## Summary of Mixed Model Analyses

**mouse average MLH1 count**

Multiple models were tested to test for potential evo models -- for the mouse average CO counts.  (there were a mix of results -- and nuanced)

First full Mixed model -- no significant fixed factors --(the random effect might be significant).

@@

**Variance effect**

** Do these patterns hold across Quality bins? **

@@

<!-- intro
in addition to the mean gwRR, the variance is also an important metric to describe

above method remove / accounts for mean-variance correlation -

The fig5 also has a histogram of within animal variances in MLH1 counts (For F2s, with the parents as reference) (I could do this for males and females)

QTL on chrm 14, decreased variance in GG genotype, compared to hetZ and WSB (I think this also raised the mean)

Previous vQTLs (Shen etal 2012, Hulse and Cai 2013, Mulder et al 2016)

meiotic Checkpoints for ensuring consistency (Hochwagen and Amon 2006, Li and Schimenti 2007, Jaramillo-Lambert 2010 and Kauppi 2013)
-->

Histograms might not be the best way to represent, but they show the general pattern than female mice have higher within mouse variation for CO counts.

A general pattern is that the female data has higher within mouse variance (both var and cV). Some strains more than others (investigate potential outliers). (LEW females with top 3 variance are all from different dissection dates)
This pattern also holds for within mouse cv for MLH1 count (not shown).

@@

```{r mouse.var.plots.female, echo=FALSE, fig.width = 4, fig.height=2.5}

Mol.female.VAR.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.female.VAR.plot

```{r mouse.var.plots.male, echo=FALSE, fig.width = 4, fig.height=2.5}

from the Mixed models, there should be 4, 1-2 models from the full dataset for var and cv, and 3-4 from a higher quality data set for var and cv. Only a subset of the model outputs will be shown.

For the mixed models the indiviual fixed effects are tested by a anova of the full and reduced models. The interaction fixed effect pvalue comes from drop1() a LTR test.

$$mouse metric  ~=~ subsp * sex + rand(strain)+ \varepsilon$$

```{r M1.MLH1.var_cv, echo=FALSE}
#modeling within mouse variance
#use  DF.HetC.MixedModel.HQ

#HQ var.CO
MLH1var_M1 <- lmer( var ~ subsp * sex +(1|strain), data=DF.HetC.MixedModel.HQ )
MLH1var_M1.reduce.sex <- lmer( var ~ subsp  +(1|strain), data=DF.HetC.MixedModel.HQ )
MLH1var_M1.reduce.subp <- lmer( var ~ sex  +(1|strain), data=DF.HetC.MixedModel.HQ )

#summary(MLH1var_M1)
#drop1(MLH1var_M1, test="Chisq") #interaction term NS
#when switched to additive fixed, sex is very sig

#anova(MLH1var_M1, MLH1var_M1.reduce.sex)
#anova(MLH1var_M1, MLH1var_M1.reduce.subp)

#exactRLRT(MLH1var_M1) #strain
#as.data.frame( round(summary(MLH1var_M1)$coefficients, 6))

#drop1(MLH1cV_M1, test="Chisq") #interaction term NS #when switched to additive fixed, sex is very sig
#anova(MLH1cV_M1, MLH1cV_M1.reduce.sex)
#anova(MLH1cV_M1, MLH1cV_M1.reduce.subp)
#exactRLRT(MLH1cV_M1) #strain

#drop1(MLH1var_M1_Q12, test="Chisq")#sig

>>>>

---
title: "MLH1 Results Outline"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

options(digits=2)

@@

```{r FEMALE.TABLE, echo=FALSE}

#Mol
#Mol.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Mol")

#Dom 
Dom.female_mouse.av <-  female_mouse.avs_for.analysis %>% filter(sex== "female") %>% filter(subsp == "Dom")

```{r MALE.TABLE, echo=FALSE}

Musc.male_mouse.av <-  fomo.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Musc")
Mol.male_mouse.av <-  fomo.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Mol")
#Dom version 
Dom.male_mouse.av <-  fomo.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Dom")
```

```{r MIXED.MODEL.setup, echo=FALSE, warning=FALSE, eval=TRUE}
#DF.HetC.MixedModel.HQ

@@

@@

#this move this to the setup file
#MLH1_data  %>% filter(quality < 5)  -- low quality cells removed

#pooled
strain.table.HQ_Fig.og <- ddply(MLH1_data  %>% filter(quality < 5), c("subsp","strain", "sex"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  =length(nMLH1.foci),
                        mean_co =as.numeric(format(round(mean(nMLH1.foci), 3 ), nsmall=3) ),
                        cV = cv(nMLH1.foci, na.rm = TRUE),
                        var = as.numeric(format(round(   var(nMLH1.foci, na.rm = TRUE),3), nsmall=3)), 
                        sd   = round(sd(nMLH1.foci, na.rm = TRUE), 3),
                       #Standard error (mice vs cells)
                        se  = round(sd / sqrt(Ncells), 3)

                       #range_mouse_means = range()
                   #consider adding crhm proportions?
    )

@@

```{r main.fig1s, echo=FALSE, include=FALSE}

@@

#strain.table.HQ_Fig.og  all strains
#strain.table.HQ_Fig  extra strains removed

main.table <- strain.table.HQ_Fig.og  %>% filter(strain != "F1")

main.table <- add_subsp(main.table)

kable(head(main.table),  caption = "Table 1. MLH1 counts per cell summary statistics by strain")

#strain.table.HQ_Fig

#sum( (main.table  %>% filter(sex == "male"))$Ncells)
#the kable version is bulkier than the other versaion
```

In order to estimate the genome wide recombination rates across our panel of `r length(unique(main.table$strain))` strains, MLH1 foci were quantified from a total of `r sum( (main.table  %>% filter(sex == "male"))$Ncells)` spermatocytes and `r sum( (main.table  %>% filter(sex == "female"))$Ncells)`.

The strain means for females were pretty similar. For female Domesticus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Dom")  )$mean_co` for _M. m. domesticus^WSB^_, _M. m. domesticus^G^_, and _M. m. domesticus$^LEW^_ respectively. In the three Musculus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Musc")  )$mean_co` for _M. m. musculus^ PWD}^_, _M. m. musculus^ SKIVE^_, and _M. m. musculus^KAZ^_ respectively. In the molossinus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Mol")  )$mean_co` for _M. m. molossinus^ MSM^_ and _M. m. molossinus^MOLF^_ respectively.

While in two of the subspecies, the means for males show a greater range across means, notable in the mmmolossinus strains `r (main.table  %>% filter(sex == "male") %>% filter(subsp == "Mol")  )$mean_co` for _M. m. molossinus^MSM^_ and  _M. m. molossinus^MOLF^_ respectively. In the musculus the strain means are `r (main.table  %>% filter(sex == "male") %>% filter(subsp == "Musc")  )$mean_co` for _M. m. musculus^PWD^_, _M. m. musculus^SKIVE^_, _M. m. musculus^KAZ^_, _M. m. musculus^TOM^_, _M. m. musculus^AST^_, and _M. m. musculus^CZECH^_ respectively. In the male Dom strains the means for MLH1 count per cell are `r (main.table  %>% filter(sex == "male") %>% filter(subsp == "Dom")  )$mean_co`  for  _M. m. domesticus^WSB^_, _M. m. domesticus^G^_, _M. m. domesticus^LEW^_ and _M. m. domesticus^PERC^_ respectively. These

<!-- mouse level means-->
The ranges of mean MLH1 counts per cell are

<!--(the figure is made from pooling MLH1 cells)
the male

- the ranges of mouse means are displayed in Sup table X

- How the means were calculated

-The male gwRR is adjusted for the PAR
-The means for each strains are report in Table 1

-Normality was assessed for the MLH1 counts (include in Supplement histograms)

, -->

@@

In order to test the effects of subspecies, sex and strain on the mean and variance of MLH1 counts per cell, we fit a data set of `r length(DF.HetC.MixedModel.HQ$mean_co )` mouse averages for MLH1 foci per cell. We constructed the model to set subspecies, sex and their interaction as fixed effects, which strain was coded as a random effect. The predicted coeffecients and pvalues are reporter below in table X.

$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

<!--
This is the break down of mice used within the Mixed model

@@

In order to understand the variance within an evolutionary framework, we fit `r length(DF.HetC.MixedModel.HQ$mean_co )` mouse level MLH1 foci per cell means as the dependant variable with a mixed model.

Setting up the data set for using the mixed model (base data set, HQ data set)

subspecies levels: 3, Musc, Dom and Molossinus
strain levels: Dom: 3, Musc: 3  Mol: 2
sex: 2 female male  
-->

@@

summary(MLH1.MM.M1)$coefficients

MLH1.MM_interaction <- drop1(MLH1.MM.M1, test="Chisq")$`Pr(Chi)`[2]
MLH1.MM_sex <- anova(MLH1.MM.M1, MLH1.MM.M1.reduce.sex)$`Pr(>Chisq)`[2]
MLH1.MM_subp <- anova(MLH1.MM.M1, MLH1.MM.M1.reduce.subp)$`Pr(>Chisq)`[2]
MLH1.MM_strain<- MM.rand.pval <- exactRLRT(MLH1.MM.M1)$p.value

MLH1.MMlmer_results <- data.frame(sub.sp = MLH1.MM_subp, 
sex.results =  MLH1.MM_sex, 
               interaction.pval = MLH1.MM_interaction,  rand.pval = 
MLH1.MM_strain)

Because all of the mixed model effects had significant effects, (p= `r  round(MLH1.MMlmer_results$sub.sp, 4)`, `r  round(MLH1.MMlmer_results$sex.results, 4)`, and  `r  round(MLH1.MMlmer_results$interaction.pval, 4)` for the sex, subspecies and interaction fixed effects and `r  MLH1.MMlmer_results$rand.pval` for the random strain effect), we followed up with two general linear models to investigate the specific strain effects.

 $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$

and

$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$

@@

summ <- summary(glm.strain.fixed.full)
#table of just coeff anf pvalue
#round(summ$coefficients[,c(1,4)],5 )

#write.table(summ$coefficients , "~./MLH1repo/glm.strain.fixed.full.csv", sep=",",row.names = TRUE)
```

In the first glm, the fixed subspecies effects were not significant, so we only report for values for the second model which is limited to sex and strain fixed effects.

@@

#write.table(summ.strain.fixed$coefficients , "~./MLH1repo/summ.strain.fixed.csv", sep=",",row.names = TRUE)
```

In both models strain by sex interaction effects were significant for (p = `r summ.strain.fixed$coefficients[,4][12]` and `r summ.strain.fixed$coefficients[,4][13]` for the second mixed model) for _M. m. musculus^PWD^_ and _M. m. molossinus^MSM^_  respectively. These results confirm the qualitative observation that within these two strains there has been sex-specific in the average MLH1 count per cell.

Additionally, _M. m. domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.strain.fixed$coefficients[,4][3]` and `r summ.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _M. m. domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

@@

@@

In order to more fully discribe the distributions we examined the variance in addition to the mean applying the same mixed models and linear models, with _within mouse variance_ of MLH1 counts per cell as the dependant variable.

<!--
motivation / background: maybe checkpoint / regulation

limitations: observations at the mouse level, decreased power to estiamate the variance --

setup

-quality / technical error

- variance and cv

The mixed models have most significant variables,

M3.glm for variance was less significant (0.044181 ),

the Q12 models in general were more significant

cV's were more significant than var

var, full.mixed, q12,mixed, M3.glm.full, M3.glm.q12
1.618e-10 ***, 2.092e-11, 0.044181 * (male),  0.000228 ***

cv, full.mixed, q12,mixed, M3.glm.full, M3.glm.q12
5.033e-12 ***, 1.315e-12,  3.195e-12 ***, 3.16e-05 ***
-->

$$mouse\ variance\ for\ MLH1\ foci\ per\ cell   ~=~ subsp * sex + rand(strain)+ \varepsilon$$

```{r M1.MLH1.var_cv, echo=FALSE, warning=FALSE, message=FALSE}

MLH1var_M1 <- lmer( var ~ subsp * sex +(1|strain), data=DF.HetC.MixedModel.HQ )
MLH1var_M1.reduce.sex <- lmer( var ~ subsp  +(1|strain), data=DF.HetC.MixedModel.HQ )
MLH1var_M1.reduce.subp <- lmer( var ~ sex  +(1|strain), data=DF.HetC.MixedModel.HQ )

#summary(MLH1var_M1)
LmerMLH1_VAR_int <- drop1(MLH1var_M1, test="Chisq")$`Pr(Chi)`[2]
LmerMLH1_VAR_sex <- anova(MLH1var_M1, MLH1var_M1.reduce.sex)$`Pr(>Chisq)`[2]
LmerMLH1_VAR_subsp <- anova(MLH1var_M1, MLH1var_M1.reduce.subp)$`Pr(>Chisq)`[2]
LmerMLH1_VAR_strain <- exactRLRT(MLH1var_M1)$p.value

LmerMLH1_VAR_results <- data.frame(sub.sp = LmerMLH1_VAR_subsp,
sex.results =  LmerMLH1_VAR_sex, interaction.pval = LmerMLH1_VAR_int,  rand.pval = LmerMLH1_VAR_strain)
LmerMLH1_VAR_results

LmerMLH1_CV_int <- drop1(MLH1cV_M1, test="Chisq")$`Pr(Chi)`[2]
LmerMLH1_CV_sex <- anova(MLH1cV_M1, MLH1cV_M1.reduce.sex)$`Pr(>Chisq)`[2]
LmerMLH1_CV_subsp <- anova(MLH1cV_M1, MLH1cV_M1.reduce.subp)$`Pr(>Chisq)`[2]
LmerMLH1_CV_strain <- exactRLRT(MLH1cV_M1)$p.value

LmerMLH1_CV_results <- data.frame(sub.sp = LmerMLH1_CV_subsp,
sex.results =  LmerMLH1_CV_sex, interaction.pval = LmerMLH1_CV_int,  rand.pval = LmerMLH1_CV_strain)
LmerMLH1_CV_results

LmerMLH1_M1_Q12VAR_int <- drop1(MLH1var_M1_Q12, test="Chisq")$`Pr(Chi)`[2]
LmerMLH1_M1_Q12_VAR_sex <- anova(MLH1var_M1_Q12, MLH1var_M1_Q12reduce.sex)$`Pr(>Chisq)`[2]
LmerMLH1_M1_Q12_VAR_subsp <- anova(MLH1var_M1_Q12, MLH1var_M1_Q12reduce.subp)$`Pr(>Chisq)`[2]
LmerMLH1_M1_Q12_VAR_strain <- exactRLRT(MLH1var_M1_Q12)$p.value

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_VAR_subsp,
sex.results =  LmerMLH1_M1_Q12_VAR_sex, interaction.pval = LmerMLH1_M1_Q12VAR_int,  rand.pval = LmerMLH1_M1_Q12_VAR_strain)
LmerMLH1_M1_Q12_VAR_results

LmerMLH1_M1_Q12_CV_int <- drop1(MLH1cv_M1_Q12, test="Chisq")$`Pr(Chi)`[2]
LmerMLH1_M1_Q12_CV_sex <- anova(MLH1cv_M1_Q12, MLH1cv_M1_Q12reduce.sex)$`Pr(>Chisq)`[2]
LmerMLH1_M1_Q12_CV_subsp <- anova(MLH1cv_M1_Q12, MLH1cv_M1_Q12reduce.subp)$`Pr(>Chisq)`[2]
LmerMLH1_M1_Q12_CV_strain <- exactRLRT(MLH1cv_M1_Q12)$p.value

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_CV_subsp,
sex.results =  LmerMLH1_M1_Q12_CV_sex, interaction.pval = LmerMLH1_M1_Q12_CV_int,  rand.pval = LmerMLH1_M1_Q12_CV_strain)

LmerMLH1_M1_Q12_VAR_results
```

These are the p values for the effects of the mixed models of the foloowing mixed models: VAR.FULL, CV.FULL, VAR.Q12 and cv.Q12

@@
